Supplementary Materials And Methods

### 2020-08-14 Supplementary_mat_met.Rmd
article current files

---
title: "Supplementary Materials And Methods"
author: 
- Selcen Ari
- Alper Yilmaz 
date: "7 Nisan 2019"
output:
  pdf_document: default
  html_document: default
  word_document: default
bibliography: ref.bib
header-includes: 
  \usepackage{float} \floatplacement{figure}{H} 
  \newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{S\arabic{table}} \setcounter{figure}{0}     \renewcommand{\thefigure}{S\arabic{figure}}}
  \usepackage{subfig}
---

@@

**Trigger change in graph**: The dataset are assumed as steady-state in previous step and the efficiency coefficients are calculated according to this acceptance. In the network that is found in steady-state conditions, the change is applied to the graph object for distribution of steady-state. To provide the distribution in the network the workflow offer two methods: update_variables and update_how. The first, a new dataset that is contained competing and repressive element names and current values of these can be processed with update_variables. The second option, the amount of the given node name in update_how function can be changed according to “how” argument.

**Updating current values of variables**: After variables updating in edge variables, these are carried to node variables. Current and previous values of variables are stored as node variables with update_variables function.

@@

The node elements in the dataset are handled as two type; repressive (miRNAs) and competing (targets). It is assumed in approach that while targets are degrading or inhibiting by miRNAs continuously, miRNAs reversibly used. If the trigger of the network is a miRNA, it maintains the current value of amount that provides by user. On the contrary, it tries to help this process to provide steady-state through the regulations on its amount, if a competing element is used as a trigger. The functions that are used in the this study are developed with R and their source code is available in [Github repository](https://github.com/selcenari/ceRNAnetsim).

@@

library(tidygraph)
library(dplyr)
library(tidyr)
library(stringr)
library(readr)
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=50),tidy=TRUE)

@@

```{r, message=FALSE, warning=FALSE}
#install.packages("devtools")
#devtools::install_github("selcenari/ceRNAnetsim")
library(ceRNAnetsim)
```

@@

```{r}
minsamp<-readRDS("data/minsamp.RDS")

minsamp
```

See Table S1 in Supplementary Tables file.

@@

When the graphs were examined, it was observed that behaviors were same. But, when the results were analysed in terms of expression values, the regulation differences can be observed.

@@

The common target perturbation  (increasing to two fold at Gene4 expression in presence of interaction factors) resulted in more prominent efficiency at the same conditions (shown in following).

@@

find_node_perturbation(sample_graph, how = 2, cycle = 3, limit = 0.1)

@@

# 6. Obtaining breast cancer dataset and integration

This section describes how to apply `ceRNAnetsim` package on a breast cancer patient miRNA:target interaction dataset. Before the approach, we obtained three datasets and combined them.

```{r, echo=FALSE, message= FALSE, warning=FALSE, eval=FALSE}
# Install the packages which is necessary to obtain and process datasets.
#install.packages("BiocManager")
#BiocManager::install("SummarizedExperiment")
#BiocManager::install("TCGAbiolinks")
library(TCGAbiolinks)
library(SummarizedExperiment)
```

## 6.1 How to get gene expression counts of TCGA-E9-A1N5 patient.

We have obtained the gene expression values of patient using [TCGAbiolinks](https://www.bioconductor.org/packages/release/bioc/html/TCGAbiolinks.html) package from [Bioconductor](https://www.bioconductor.org/). For this process, we have followed the instructions of the package. `TCGAbiolinks` package provides to obtain data for whole number of given barcode(s) at once. But, we preferred to download them separately to show datasets.

+ Obtain to gene expression counts of tumor tissue.

```{r,message= FALSE, warning=FALSE, eval=FALSE}

BCP_tumor <- GDCquery(project = "TCGA-BRCA",
                  data.category = "Transcriptome Profiling",
                  data.type = "Gene Expression Quantification", 
                  workflow.type = "HTSeq - Counts",
                  barcode = "TCGA-E9-A1N5-01A-11R-A14D-07")

GDCdownload(BCP_tumor)
BCPGE_tumor <- GDCprepare(BCP_tumor)

as.data.frame(assay(BCPGE_tumor))%>%
  mutate(ensembl_gene_id = rownames(.))%>%
  dplyr::inner_join(as.data.frame(rowData(BCPGE_tumor)), by = "ensembl_gene_id")%>%
  dplyr::select(ensembl_gene_id, external_gene_name, 1)-> TCGA_E9_A1N5_tumor

colnames(TCGA_E9_A1N5_tumor)[3] <- "GE_tumor"

head(TCGA_E9_A1N5_tumor)
```

+ Obtain to gene expression counts of normal tissue.

```{r, message= FALSE, warning=FALSE, eval=FALSE}
BCP_normal <- GDCquery(project = "TCGA-BRCA",
                  data.category = "Transcriptome Profiling",
                  data.type = "Gene Expression Quantification", 
                  workflow.type = "HTSeq - Counts",
                  barcode = "TCGA-E9-A1N5-11A-41R-A14D-07")
GDCdownload(BCP_normal)

BCPGE_normal <- GDCprepare(BCP_normal)

as.data.frame(assay(BCPGE_normal))%>%
  mutate(ensembl_gene_id = rownames(.))%>%
  dplyr::inner_join(as.data.frame(rowData(BCPGE_normal)), by = "ensembl_gene_id")%>%
  dplyr::select(ensembl_gene_id, external_gene_name, 1)-> TCGA_E9_A1N5_normal

colnames(TCGA_E9_A1N5_normal)[3] <- "GE_normal"

head(TCGA_E9_A1N5_normal)

@@

## 6.2 How to get miRNA expression counts of TCGA-E9-A1N5 patient.

We have used `TCGAbiolinks` package to obtain miRNA expression quantification. The query gives read count of miRNA as isoform chromosome coordination. The data also contains mature miRNA information. So, we processed data to attain -5p -3p isoform information using mirbase release21 dataset.

+ Get the mirbase id of mature miRNA:

We downloaded the mirbase release 21 dataset from [mirbase](http://www.mirbase.org/ftp.shtml) and processed the patient miRNA expression datasets as following:

```{r, message= FALSE, warning=FALSE}

read_tsv("hsa_mirna.txt", comment = "#", col_names = FALSE)%>%
  dplyr::select(mirna_type= X3, definition = X9)%>%
  filter(!endsWith(mirna_type, "primary_transcript"))%>%
  tidyr::separate(definition, c("ID", "Alias", "Name", "Derivated"), sep = ";")%>%
  dplyr::select(Alias, Name)%>%
  tidyr::separate(Alias, c("trash1", "ID"), sep = "=")%>%
  tidyr::separate(Name, c("trash2", "Name"), sep = "=")%>%
  dplyr::select(-trash1, -trash2)-> mirbase_id_conv

head(mirbase_id_conv)
```

+ Obtain the miRNA expression of tumor tissue of patient:

```{r, message= FALSE, warning=FALSE, eval=FALSE}
BCP_mirnatumor <- GDCquery(project = "TCGA-BRCA",
                  data.category = "Transcriptome Profiling",
                  data.type = "Isoform Expression Quantification",
                  workflow.type = "BCGSC miRNA Profiling",
                  barcode = "TCGA-E9-A1N5-01A-11R-A14C-13")

GDCdownload(BCP_mirnatumor)

GDCprepare(BCP_mirnatumor)%>%
  as.data.frame()%>%
  dplyr::select(miRNA_ID, read_count, reads_per_million_miRNA_mapped, miRNA_region)%>%
  dplyr::filter(startsWith(miRNA_region, "mature"))%>%
  dplyr::mutate(mirbase_id =str_remove(miRNA_region, "mature,"))%>%
  dplyr::select(-miRNA_region)%>%
  dplyr::inner_join(mirbase_id_conv, by = c("mirbase_id"="ID"))%>%
  dplyr::select(miRNA_name = Name, read_count, reads_per_million_miRNA_mapped)%>%
  dplyr::group_by(miRNA_name)%>%
  mutate(read_count= sum(read_count), reads_per_million_miRNA_mapped = sum(reads_per_million_miRNA_mapped))%>%
  dplyr::ungroup()%>%
  distinct() -> BCPME_mirnatumor

head(BCPME_mirnatumor)
```

+ Obtain the miRNA expression of normal tissue of patient:

```{r, message= FALSE, warning=FALSE, eval=FALSE}
BCP_mirnanormal <- GDCquery(project = "TCGA-BRCA",
                  data.category = "Transcriptome Profiling",
                  data.type = "Isoform Expression Quantification",
                  workflow.type = "BCGSC miRNA Profiling",
                  barcode = "TCGA-E9-A1N5-11A-41R-A14C-13")

GDCdownload(BCP_mirnanormal)
# a616435d-0b69-48ac-813d-5d75ad9b85eb.mirbase21.isoforms.quantification.txt

GDCprepare(BCP_mirnanormal)%>%
  as.data.frame()%>%
  dplyr::select(miRNA_ID, read_count, reads_per_million_miRNA_mapped, miRNA_region)%>%
  dplyr::filter(startsWith(miRNA_region, "mature"))%>%
  dplyr::mutate(mirbase_id =str_remove(miRNA_region, "mature,"))%>%
  dplyr::select(-miRNA_region)%>%
  dplyr::inner_join(mirbase_id_conv, by = c("mirbase_id"="ID"))%>%
  dplyr::select(miRNA_name = Name, read_count, reads_per_million_miRNA_mapped)%>%
  dplyr::group_by(miRNA_name)%>%
  mutate(read_count= sum(read_count), reads_per_million_miRNA_mapped = sum(reads_per_million_miRNA_mapped))%>%
  dplyr::ungroup()%>%
  distinct() -> BCPME_mirnanormal

head(BCPME_mirnanormal)
```

## 6.3 Get the high-throughput experimental miRNA:target dataset.

There are various datasets about miRNA:target pairs such as miRTarBase, DianaTools, miRecords, miRWalk etc. Some of these present the experimentally supported miRNA target pairs or only predicted ones. The experimentally supported datasets generally provides weak evidence for interactions. For these reasons, we obtained the high-throughput experimental miRNA:target dataset from two studies performed by Helwak et al. and Moore et al. These steps were not handle in this file because they contain many processes.

Briefly these datasets contain various common information about miRNA:target interactions such as the miRNA name, miRNAsequence, target name, target sequence, their chromosomal locations, binding location on the target sequence, binding free energy, seed structure. But these datasets provides the informations with different data structures.So we followed the steps:

+ The datasets were directly downloaded from supplementary data files of the studies.
+ It was provided that the datasets are converted to same human genome build.
+ The seed type information was organized as the same style.
+ The datasets were combined.
+ We committed the interaction factors as numeric values according to previous studies. (We added the interaction factors and their numeric values. The codes about integration of high-throughput experimental studies are available at Supplementary additional file and numeric values of interaction factors can be found at Table S2-3 in Supplementary Tables file.)

Finally, we have obtained the experimentally supported miRNA:target dataset.

```{r, message= FALSE, warning=FALSE}

experimentalmirnagene<-readRDS("data/experimentalmirnagene.RDS")

head(experimentalmirnagene)

@@

The methods about miRNA:target interactions are based a basic principle that is reading after isolation of miRNA:target chimerics. The datasets contain all the chimeric miRNA:target structures found in the medium during the experiment. On the other hand, it could be said that the reading is performed as snapshot.Because of that, the methods can provide different chimeric interactions the same miRNA:target pair. We have preferred to select most effective interaction parameters for the same miRNA:target pairs that can exhibit various interactions. The step is performed as:

```{r, message= FALSE, warning=FALSE, eval=FALSE}
experimentalmirnagene%>%
  dplyr::select(miRNA, Ensembl_Gene_Id, hgnc_symbol, Energy, seed_type_effect, region_effect)%>%
  distinct()%>%
  group_by(Ensembl_Gene_Id, miRNA)%>%
  mutate(seed_type_effect= ifelse(seed_type_effect==max(seed_type_effect), seed_type_effect, max(seed_type_effect)), Energy = ifelse(Energy==min(Energy), Energy, min(Energy)), region_effect= ifelse(region_effect==max(region_effect), region_effect, max(region_effect)))%>%
  distinct()-> tocombine_mirnagene

head(tocombine_mirnagene)
```

## 6.4 Combine the dataset

```{r, message= FALSE, warning=FALSE, eval=FALSE}
BCPME_mirnanormal%>%
  dplyr::inner_join(tocombine_mirnagene, by = c("miRNA_name"="miRNA"))%>%
  dplyr::inner_join(TCGA_E9_A1N5_normal, by = c("Ensembl_Gene_Id"="ensembl_gene_id", "hgnc_symbol"="external_gene_name"))%>%
  distinct()%>%
  dplyr::select(hgnc_symbol, miRNA_name, mirna_RPM= reads_per_million_miRNA_mapped, GE_normal, Energy, seed_type_effect, region_effect)-> E9GE_mirnagenenormal

@@

A part of dataset can be found at Table S4 in Supplementary Tables.

```{r}
E9GE_mirnagenenormal<-readRDS("data/E9GE_mirnagenenormal.RDS")
head(E9GE_mirnagenenormal)
```

```{r, message= FALSE, warning=FALSE, eval=FALSE}
BCPME_mirnatumor%>%
  dplyr::inner_join(tocombine_mirnagene, by = c("miRNA_name"="miRNA"))%>%
  dplyr::inner_join(TCGA_E9_A1N5_tumor, by = c("Ensembl_Gene_Id"="ensembl_gene_id", "hgnc_symbol"="external_gene_name"))%>%
  distinct()%>%
  dplyr::select(hgnc_symbol, miRNA_name, mirna_RPM= reads_per_million_miRNA_mapped, GE_tumor, Energy, seed_type_effect, region_effect)-> E9GE_mirnagenetumor

@@

```{r}
#saveRDS(E9GE_mirnagenetumor, "data/E9GE_mirnagenetumor.RDS")

E9GE_mirnagenetumor<-readRDS("data/E9GE_mirnagenetumor.RDS")
head(E9GE_mirnagenetumor)
```

## 6.5 Selection of trigger node

We have determined the most important nodes of network. We applied `find_node_perturbation` function. We only defined nodes that affect the other nodes more than 1.05 fold change with 10 iteration when they increase 3 fold.

```{r, include=FALSE, eval=FALSE}

as_data_frame(E9GE_mirnagenenormal)%>%
  priming_graph(competing_count = GE_normal, miRNA_count = mirna_RPM, aff_factor = c(Energy, seed_type_effect), deg_factor = region_effect)%>%
  find_node_perturbation(how = 3, cycle = 10, fast = 5)%>%
  filter(!is.na(perturbation_efficiency))->perturbationofnetwork

perturbationofnetwork%>%
  filter(type== "Competing")%>%
  arrange(desc(perturbation_efficiency))
```
This dataset, `perturbationofnetwork`, includes totally 423 effective nodes. We selected SERPINE2 gene to perturb the network because it is the most efficient node gene in network.

## 6.6 Approach of Method into Combined Datasets

We selected SERPINE2 gene for simulation of regulation on network.

### 6.6.1 Find iteration of simulation

The node amount of changed gene on the system in terms of percentage were shown at Figure S9 in Supplementary Figures. As seen, firstly, the changed gene count increase. The system which contains the hundreds of miRNAs and thousands of genes can slowly gain the steady-state again. At first glance, it can be assumed that when all nodes in the system are reached, stable state will be provided. However, although all nodes are reached, the nodes competing with each other cause the edits to continue for a while.

The dynamics of the approach are shown in package  [vignettes](https://selcenari.github.io/ceRNAnetsim/articles/convenient_iteration.html).

So, we offered an approach about to find iteration. `find_iteration` function does not give the iteration to gain steady-state, but it gives the iteration which has maximum affected node counts . The function is applied as following:

```{r}

as.data.frame(E9GE_mirnagenenormal)%>%
 priming_graph(competing_count = GE_normal, miRNA_count = mirna_RPM, aff_factor = c(Energy, seed_type_effect), deg_factor = region_effect)%>%
 update_how("SERPINE2",2.75)%>%
 simulate(50) %>%
 find_iteration(limit=1, plot= FALSE)

@@

### 6.6.2 Simulation of dataset

We tried to apply two fold of the point that SERPINE2 has maximum affected genes on network.

```{r}
as.data.frame(E9GE_mirnagenenormal)%>%
  priming_graph(competing_count = GE_normal, miRNA_count = mirna_RPM, aff_factor = c(Energy, seed_type_effect), deg_factor = region_effect)%>%
  update_how("SERPINE2",2.75)%>%
  simulate(62)

@@

```{r}
as.data.frame(E9GE_mirnagenenormal)%>%
  priming_graph(competing_count = GE_normal, miRNA_count = mirna_RPM, aff_factor = c(Energy, seed_type_effect), deg_factor = region_effect)%>%
  update_how("SERPINE2",2.75)%>%
  simulate(62)%>%
  as_tibble()%>%
  dplyr::select(name, initial_count, count_current)->simulation_results
```

### 6.6.3 Comparison of simulation results and tumor tissue expression values
```{r}
E9GE_mirnagenetumor%>%
  dplyr::select(hgnc_symbol, GE_tumor)%>%
  dplyr::inner_join((E9GE_mirnagenenormal%>%dplyr::select(hgnc_symbol, GE_normal)), by ="hgnc_symbol")%>%
  inner_join(simulation_results, by= c("hgnc_symbol"="name"))%>%
  distinct()
```

Actually, we have developed to provide a new approach miRNA mediated regulation networks. This approach may not explain the whole regulation behaviors between miRNAs and targets but can be first step to more detailed and coherent miRNA:target regulation approach.

>>>>

---
title: "Supplementary Materials And Methods"
author: 
- Selcen Ari
- Alper Yilmaz 
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
  word_document: default
bibliography: NAR-sample.bib
header-includes: 
  \usepackage{float} \floatplacement{figure}{H} 
  \newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{S\arabic{table}} \setcounter{figure}{0}     \renewcommand{\thefigure}{S\arabic{figure}}}
  \usepackage{subfig}
---
\tableofcontents

@@

**Trigger change in graph**: The dataset are assumed as steady-state in previous step and the efficiency coefficients are calculated according to this acceptance. In the network that is found in steady-state conditions, the change is applied to the graph object for perturbation of steady-state. To provide the perturbation in the network the work-flow offer two methods: *update_variables* and *update_how*. The first, a new dataset that is contained competing and repressive element names and current values of these can be processed with update_variables. The second option, the amount of the given node name in update_how function can be changed according to “how” argument.

@@

The node elements in the dataset are handled as two type; repressive (miRNAs) and competing (targets). It is assumed in approach that while targets are degrading or inhibiting by miRNAs continuously, miRNAs reversibly used. If the trigger of the network is a miRNA, it maintains the current value of amount that provides by user. On the contrary, it tries to help this process to provide steady-state through the regulations on its amount, if a competing element is used as a trigger. The functions that are used in the this study are developed with R and are available in [Bioconductor](https://www.bioconductor.org/packages/release/bioc/html/ceRNAnetsim.html).

@@

library(tidygraph)
library(dplyr)
library(tidyr)
library(stringr)
library(readr)
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=50),tidy=TRUE)
knitr::opts_chunk$set(warning = FALSE)

@@

```{r, message=FALSE, warning=FALSE}
library(ceRNAnetsim)
```

@@

```{r}
data("minsamp")
minsamp
```

See Table S1 in Supplementary Figures and Tables file.

@@

When the graphs were examined, it was observed that behaviors (i.e. fluctuating trend) were same. But, when the results were analysed in terms of expression values, the regulation differences can be observed.

@@

The common target perturbation  (increasing to two fold at Gene4 expression in presence of interaction factors) resulted in more prominent efficiency at the same conditions (shown at following).

@@

find_node_perturbation(sample_graph, how = 2, cycle = 3, limit = 0.1)%>%
  as_tibble()%>%
  select(name, perturbation_efficiency, perturbed_count)

@@

# 6. Additional data manipulation steps

## 6.1 Arrangement of CLASH dataset

CLASH dataset was retrieved from [PubMed](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3650559/) [@helwak_mapping_2013].

```{r, eval=FALSE}

clashelwak <- read.table("mmc1.txt", comment.char = "#", header = TRUE, skip = 1, stringsAsFactors = FALSE)

#hg19
```

### Query of Human Genome 19.

Human genome 19 information was handled through biomaRt package.

```{r,  eval=FALSE}
#HG19
listEnsemblArchives()
listMarts(host = 'http://grch37.ensembl.org')
ensemblgrch37 = useMart(host='http://grch37.ensembl.org', 
                     biomart='ENSEMBL_MART_ENSEMBL',
                    dataset='hsapiens_gene_ensembl')
hg19 <- getBM(attributes = c("ensembl_transcript_id","ensembl_gene_id","chromosome_name", "start_position","end_position", "hgnc_symbol", "entrezgene_id", "strand" ), mart = ensemblgrch37)

@@

### Adding miRNA and gene information

```{r,  eval=FALSE}

clashelwak <- clashelwak%>%
 separate(microRNA_name, c("Barcode", "Database", "mirna_name", "type"), sep = "_")%>%
  separate(mRNA_name, c("Ensembl_Gene_Id", "Ensembl_Transcript_Id", "Hugo_Symbol", "mRNA_Type"), sep = "_")

@@

MiRNA releases are obtained from [miRBase](http://www.mirbase.org/ftp.shtml). In this step, release 21 (in Human genome 38) was downloaded.

```{r, eval=FALSE}
read.table("mirbasehg38.txt", comment.char = "#")%>%
  filter(V3 != "miRNA_primary_transcript")%>%
  separate(V9, c("ID", "Alias", "Name", "Precusor"), sep = ";")%>%
  mutate(ID = substr(ID, 4, length(ID)), Alias = substr(Alias, 7, length(Alias)), Name = substr(Name, 6, length(Name)), Precusor = substr(Precusor, 14, length(Precusor)))%>%
  dplyr::select(chr= V1, start = V4, end = V5, strand = V7, ID, Alias, Name, Precusor)->mirbasehg38

@@

CLASH dataset is published in miRBase release 15 and Human Genome 19 version.

```{r, eval=FALSE}
read_tsv("mirna_mature.txt", col_names = FALSE)%>%
  filter(startsWith(X2, "hsa"))%>%
  dplyr::select(mirna_ID = X2, mirbase_ID = X3)%>%
  inner_join(mirbasehg38%>% dplyr::select(ID, Name), by= c("mirbase_ID"= "ID"))%>%
  dplyr::select(mirbase_ID, Name)%>%
  distinct()%>%
  inner_join(clashelwak, by = c("mirbase_ID"= "Barcode"))%>%
  dplyr::select(Name, miRNA_seq, Ensembl_Gene_Id, Ensembl_Transcript_Id, Hugo_Symbol, mRNA_seq_extended, chimeras_decompressed, seed_type, seed_basepairs,folding_class, seq_ID, folding_energy, X5.UTR, CDS, X3.UTR)%>%
  inner_join(hg19, by = c("Ensembl_Gene_Id"= "ensembl_gene_id", "Ensembl_Transcript_Id"="ensembl_transcript_id", "Hugo_Symbol"="hgnc_symbol"))%>%
  mutate(region1 = ifelse(X5.UTR == "1", "5UTR", " "), region2= ifelse(X3.UTR == "1", "3UTR", " "),region3= ifelse(CDS =="1", "CDS", " "))%>%
  unite(region, c(region1, region2, region3), sep = "||")%>%
  dplyr::select(chromosome_name, start_position, end_position,strand, Hugo_Symbol, Ensembl_Gene_Id, Ensembl_Transcript_Id, mRNA_seq_extended, Name, miRNA_seq, seq_ID, seed_type,seed_basepairs, folding_class, folding_energy, region)%>%
  as_tibble() -> clashelwakfinal

@@

### Converting CLASH data to human genome 38 build.

There are different liftover methods for conversion among Human Genome builds. We preffered to use [UCSC liftover tool](https://genome.ucsc.edu/cgi-bin/hgLiftOver)

```{r,  eval=FALSE}

#Obtaining chromosomal locations from miRNA:target interaction dataset.

lift19 <- clashelwakfinal%>%
  dplyr::select(1,2,3)%>%
  unite(start_end, c("start_position", "end_position"), sep = "-")%>%
  mutate(Chromosome = paste0("chr",chromosome_name, ""))%>%
  unite(chromosome_name, c("Chromosome","start_end"), sep = ":")

write_tsv(lift19, "lift19.txt")

#After we searched this file in UCSC browser, the output loaded (lift19_del deleted regions on the HG38 genome build; hg38clashcomp new locations of the genes on HG38)

lift19_del <- read_tsv("deleted_lift19.txt")
colnames(lift19_del)[1] <- "chromosome_loc"

lift19_del <- lift19_del%>%
  dplyr::filter(startsWith(chromosome_loc, "chr"))%>%
  separate(chromosome_loc, c("Chr", "End"), "-", remove = TRUE)%>%
  separate(Chr, c("Chr", "Start"), ":", remove = TRUE)

lift19_del$Start <- as.numeric(lift19_del$Start)
lift19_del$End <- as.numeric(lift19_del$End)

#removing deleted location from CLASH dataset

clashelwakfinal <-clashelwakfinal%>%
  mutate(Chromosome = paste0("chr",chromosome_name, ""))%>%
  dplyr::anti_join(lift19_del, by = c("Chromosome"="Chr", "start_position"="Start", "end_position"="End"))

hg38clash <- read.delim("hg38clashcomp.txt", header = FALSE, stringsAsFactors = FALSE)

# adding new location information:

clashelwakfinal <- clashelwakfinal%>% 
  bind_cols(hg38clash)

colnames(clashelwakfinal)[18] <- "HG38build_loc"

clashelwakfinal <- clashelwakfinal%>%
  dplyr::mutate(Genom_build= rep("hg19"))

# Arrangement in dataset

clashelwakfinal%>% dplyr::select(cluster=seq_ID, chromosome = Chromosome, start_position, end_position, strand, hgnc_symbol = Hugo_Symbol, Ensembl_Gene_Id, Ensembl_Transcript_Id, target_seq = mRNA_seq_extended, miRNA= Name, miR_seq =miRNA_seq, seed_type, seed_type2= seed_basepairs, seed_type3= folding_class, Energy = folding_energy, HG38build_loc, Genom_build, region) -> clashelwakfinal

clashelwakfinal$strand <- as.character(clashelwakfinal$strand)

str(clashelwakfinal)
```

### Interpreting the CLASH seed structures in dataset

```{r,  eval=FALSE}
clashelwakfinal%>%
  mutate(seed_type= ifelse (seed_type == "noncanonical_seed"& seed_type2 >4 &seed_type3=="I", paste0(seed_type2, "-mer"), seed_type),
         seed_type= ifelse (seed_type == "noncanonical_seed"& seed_type2 >4 &seed_type3=="II", paste0(seed_type2, "-mer_noncanonical"), seed_type), 
         seed_type= ifelse (seed_type == "noncanonical_seed"& seed_type2 >4 &seed_type3=="III", paste0(seed_type2, "-mer_noncanonical"), seed_type),
         seed_type= ifelse (seed_type == "noncanonical_seed"& seed_type2 >4 &seed_type3=="IV", paste0(seed_type2, "-mer_noncanonical"), seed_type),
         seed_type= ifelse (startsWith(seed_type, "no") ,"none", seed_type))%>%
  dplyr::select(-seed_type2, -seed_type3)->clashelwakfinal

@@

## 6.2 Arrangement of CLEAR-CLiP Dataset [@moore_mirnatarget_2015]

CLASH dataset was retrieved from [Nature article](nature.com/articles/ncomms9864)

```{r,  eval=FALSE}
clearclip <- read_xlsx("CLEAR-CLIP.xlsx")

#Clearclip hg18
```

### Query of Human Genome 18

```{r,  eval=FALSE}
#HG18
listEnsemblArchives()
listMarts(host = 'may2009.archive.ensembl.org')
ensembl54 = useMart(host='may2009.archive.ensembl.org', 
                     biomart='ENSEMBL_MART_ENSEMBL',
                    dataset='hsapiens_gene_ensembl')

hg18 <- getBM(attributes = c("ensembl_transcript_id","ensembl_gene_id","chromosome_name", "start_position","end_position", "hgnc_symbol", "entrezgene", "strand" ), mart = ensembl54)
```

### Adding Genome Information to dataset

```{r,  eval=FALSE}
clearclipfinal <- hg18%>%
  inner_join(clearclip, by= c("entrezgene"= "gene.id", "hgnc_symbol"= "gene.symbol"))%>%
  distinct()
```

### Converting human genome build

```{r,   eval=FALSE}

#Obtaining chromosomal locations from miRNA:target interaction dataset.

lift18 <- clearclipfinal%>%
  unite(start_end, c("start_position", "end_position"), sep = "-")%>%
  unite(location, c("chr","start_end"), sep = ":")%>%
  dplyr::select(location)

write_tsv(lift18, "lift18.txt")

#After we searched this file in UCSC browser, the output loaded (deleted_lift18 deleted regions on the HG38 genome build; hg38clearclip new locations of the genes on HG38)

deleted_lift18 <- read_tsv("deleted_lift18.txt")

colnames(deleted_lift18)[1] <- "Chromosome_loc"

deleted_lift18 <- deleted_lift18%>%
  dplyr::filter(startsWith(Chromosome_loc, "chr"))%>%
  separate(Chromosome_loc, c("Chr", "End"), "-", remove = TRUE)%>%
  separate(Chr, c("Chr", "Start"), ":", remove = TRUE)

deleted_lift18$Start <- as.numeric(deleted_lift18$Start)
deleted_lift18$End <- as.numeric(deleted_lift18$End)

#removing deleted location from CLEAR-CLiP dataset

clearclipfinal <- clearclipfinal%>%
  dplyr::anti_join(deleted_lift18, by = c("chr"="Chr", "start_position"="Start", "end_position"="End"))

hg38clearclip<- read.delim("hg38clearclip.txt", header = FALSE, stringsAsFactors = FALSE)

clearclipfinal <- clearclipfinal%>% 
  bind_cols(hg38clearclip)

colnames(clearclipfinal)[28] <- "HG38build_loc"

# adding new location information:

clearclipfinal <- clearclipfinal%>%
  dplyr::mutate(Genom_build= rep("hg18"))

# Arrangement in dataset

clearclipfinal%>%
  dplyr::select(cluster=cluster.ID, chromosome = chr, start_position, end_position, strand = strand.y, hgnc_symbol, Ensembl_Gene_Id =ensembl_gene_id, Ensembl_Transcript_Id =ensembl_transcript_id, target_seq = target.map, miRNA, miR_seq = miR.map, seed_type = 'seed match', Energy = MFE, HG38build_loc, Genom_build, region) -> clearclipfinal

@@

### Seed type manipulation in CLEAR-CLiP dataset

In CLEAR-CLiP dataset, seed types were shown in detail. We adjusted as canonical and non-canonical.

```{r,  eval=FALSE}
data_frame(seed_type= c("5mer_1","5mer_2", "5mer_3", "6mer", "6mer.indel", "6mer.mm", "6mer_off.mm", "6merA1", "6merA1.indel", "6merA1.mm", "7merA1", "7merA1.indel", "7merA1.mm", "7merm8", "7merm8.indel", "7merm8,mm", "8mer", "8mer.indel", "8mer.mm", "NA" ),
           seed_type_com= c("5-mer", "5-mer_noncanonical", "5-mer_noncanonical", "6-mer", "6-mer_noncanonical", "6-mer_noncanonical", "6-mer_noncanonical", "6-merA1", "6-merA1_noncanonical","6-merA1_noncanonical", "7-merA1", "7-merA1_noncanonical","7-merA1_noncanonical", "7-mer-8m", "7-mer-8m_noncanonical", "7-mer-8m_noncanonical", "8-mer", "8-mer_noncanonical", "8-mer_noncanonical", "none" )) ->clipdata_seed

clearclipfinal%>%
  inner_join(clipdata_seed, by = "seed_type")%>%
  dplyr::select(1:11, seed_type = seed_type_com, Energy, HG38build_loc, Genom_build, region)-> clearclipfinal

## 6.3 Integration of two experimental dataset

```{r,  eval=FALSE}

bind_rows(clashelwakfinal, clearclipfinal)%>%distinct() -> experimentalmirnagene

### Adding Coefficients of Interaction factors

Energy values in miRNA:target pairs are represented by high-throughput studies [@helwak_mapping_2013; @moore_mirnatarget_2015] which are utilized in this study. On the other hand, we have specified the other interaction factors, seed type and location of binding region on the target, as numeric values based on the previous studies.[@grimson_microrna_2007] have compared the seed types' effect on target repression with few miRNA had canonical seed pairing in their study. Additionally, [@bartel_micrornas:_2009] and [@betel2010comprehensive] have studied on functional and non-functional seed interactions. Based on results of these studies we have arranged seed types of miRNA:target interactions as numeric values. We also have redefined location of binding region on the target as numeric values, based on studies of [@hausser_analysis_2013] and [@helwak_mapping_2013]. With this process, we have handled this integrated dataset in context of competitor behaviors and functionality of interactions.

In this step we added numeric intraction values at followings

Fistly, we organized these values due to the fact that the regions were defined differently in two datasets. After that, region effect was added as numeric values (shown in Table S3).

```{r,  eval=FALSE}
experimentalmirnagene <- experimentalmirnagene%>%
  mutate(region2  = str_replace_all(region, "NA", ""), region3 = str_replace_all(region2, "\\|", ""), region = str_replace_all(region3, c("3'UTR"="3UTR", "5'UTR"="5UTR")))%>%
  dplyr::select(-region2, -region3)%>%
  mutate(region_effect = as.double(ifelse(region %in% c("3UTRCDS", "CDS3UTR", "5UTR3UTR", "CDS5UTR3UTR", "CDS3UTRintron"), "0.93", ifelse(region %in% c("CDS", "CDSintron"), "0.42", ifelse(region %in% c("3UTR","3UTRintron"), "0.84", ifelse(region %in% c("5UTR", "5UTRintron"), "0.01", ifelse(region %in% c("5UTRCDS", "CDS5UTR"), "0.42", ifelse(region %in% c("intron",""), "0.01", ifelse(region %in% c("exon_unclassified", ""), "0.2", NA)))))))))

Secondly, we organized seed type interactions in *Seed type manipulation* section for CLEAR-CLiP dataset to show as found in CLASH dataset. Same type formatted values added dataset as numeric values (shown in Table S2).

```{r,  eval=FALSE}

seed_type_effect <- data_frame( seed_type = c("5-mer", "5-mer_noncanonical", "6-mer", "6-mer_noncanonical", "6-merA1", "6-merA1_noncanonical", "7-mer", "7-mer_noncanonical", "7-merA1", "7-merA1_noncanonical", "7-mer-8m", "7-mer-8m_noncanonical", "8-mer", "8-mer_noncanonical", "9-mer", "9-mer_noncanonical", "none"),
                                seed_type_effect= c(0.05, 0.04, 0.07, 0.05, 0.07, 0.05, 0.23, 0.19, 0.19, 0.16, 0.25, 0.21, 0.43, 0.35, 0.43,0.35, 0.01)
                                )

experimentalmirnagene%>%
  inner_join(seed_type_effect, by= "seed_type") -> experimentalmirnagene
```

```{r}
readRDS("data/experimentalmirnagene.RDS")->experimentalmirnagene
experimentalmirnagene
```

The context of dataset is shown in Table S5 in Supplementary Tables.

# REFERENCES
